---
title: "HW 6"
author: "Lance Ding"
date: "March 2023"
output: html_document
---

```{r setup, include=FALSE}
pacman::p_load(knitr, tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions

**Write this homework acting as if I don't know what I asked you. For example, don't simply list question numbers for the headings. If you gave this document to someone else who didn't know the assignment, they should be able to understand what you did by reading the headings, code, and accompanying text.**

**Look to my HW1 and RMarkdown Organization examples for how to write good headings and organize your assignment.**

This HW is worth 10 total points. It leans heavily on activities from the [Adventures in R Course created by Dr. Kelly Bodwin of California Polytechnic State University](https://www.adventures-in-r.com/).

1. Change the author and date fields in the header above to your name and the date.

2. Make sure to load any packages you may need right at the start. Do *NOT* include the `learnr` package, ever, unless you are writing an interactive Tutorial (which you won't do in this) - this will cause problems.

3. Ensure that no chunks have the `include = FALSE` or `echo = FALSE` option, as I want to be able to see *all* your code and output.

4. Brief but descriptive headings and document organization (answers under headings, text near relevant code, brief explanatory text as indicated below, etc.) (1 pt)

5. For practice with `stringr` as well as more practice with data cleaning in a real world example, I'd like you to work through homework problems 1-4 from Dr. Martin van der Linden's HW5 assignment, which is posted on Canvas alongside this. The data `raw_ab.csv` is also posted alongside the assignment. (4.5 pts)

    * Q2 is a bit tricky because it deals with parentheses, which (like `.`) are special characters in regex. SEE *R for Data Science* section 14.3.1 for a reminder on how to deal with such characters.


6. For more practice with regular expressions, let's decode the secret message in [Puzzle 4.1](https://www.adventures-in-r.com/post/04-functions-regex/) from Adventures in R. (4.5 pts)

    HINTS AND CLARIFICATIONS:
    
    * You should do Warm-Up Exercises 1, 3, 4, and 5.
    
    * Warm-up 4: to identify and highlight any punctuation, you may use the regex character class `"[:punct:]"`
    
    * Warm-up 5: First identify (using code, not manually!) the max length of an element of the scrambled message. Then recall (or perhaps learn for the first time) you can use the notation `<VECTOR>[<LOGICAL CRITERION>]` to pull out only the elements of a vector that meet the stated criterion. For example, if you have a vector `x <- c(1:10)`, what does `x[x %% 2 == 0]` return? Try it!
    
    * Question 2: try `str_sub()`
    
    * Question 6: Unfortunately for some reason I think the message doesn't actually include anything that ends with b plus a punctuation mark. But I still want you to attempt this question as written. To check your work, feel free to simply create a separate length-1 character vector of your own that ends in b with a punctuation mark and see if you can write code to make the indicated change.
    
        Now, this one is trickier than it looks at first, but you can do it. For example, if you had the word "dumb!" this wants you to change it to "dumy!" But if you have "dumb." it wants you to change it to "dumy." Hard-coding options for each possible punctuation mark is not something I want you to do. Instead, review the section on grouping and backreferences in *R for Data Science* Ch. 14.3.5. Can you figure out how you could use this in your `replacement = ` argument to specify that you want it to replace the b with the y but keep any of the subsequent punctuation? You'll need to tell it to *refer back* to the punctuation in your `pattern =` argument.
    
    * Question 7: Don't worry about replacing with the correct case (that is, you need to identify instances of both "k" and "K", but just replace them all with either "v" or "V" - don't worry about matching them up to the case of the original k).
    
    * Don't forget to print the decoded message in an easily human-readable way using `str_c()` and tell me what movie it's from!


7. OPTIONAL OPTIONAL OPTIONAL. THIS IS NOT REQUIRED FOR THE ASSIGNMENT. ONLY DO IT IF YOU'RE CRAVING MORE PRACTICE.

For yet more practice with `stringr`, let's figure out some patterns in Billboard Hot 100 songs. Answer questions 6 and 7 from [Lab 3](https://www.adventures-in-r.com/post/03-tidy-factors/) from Adventures in R. Ignore anything in the Instructions that counters how I tell you to structure a document.

    HINTS AND CLARIFICATIONS:
    
    * Question 6: the basic path you should take to solve it is as follows. Pull out the column `title` into a character vector either using `pull()` or the `$` operator. You can then *extract all* the individual words from each song's title by using regular expressions and a `stringr` function that looks for a pattern of any number of consecutive letters (capital or lower-case) in a row. EXTRA HINT: `"[a-c]"` would look for any single instance of the lower case letters a, b, or c. You need to allow for any positive number of instances of any letter a to z, lower case or capital.
    
        If you've done this correctly, this should produce a list where each element is a character vector with a length of the number of words in that song's title. You can then use the `unlist()` function to turn this list into a character vector itself, with length equal to the total length of all its constituent vectors.
    
        You should then take *this* vector, and turn it into a `tibble()` with one column (the words themselves - we could call that column `word`). Finally, you should figure out a way to identify if a word is 4 letters or longer, then figure out a way to count up the number of times each word occurs in `word`. You can print the first however many highest values you'd like.
        
        If you've done this correctly, the second-highest word should be "your" with 955 instances.
    
    * Question 7: here you should probably begin by splitting the `artist` column into a main and featured section. Then you need to figure out a way to *split* the resulting string of featured artist(s) for each song into individual artists - note these can be split by `,`, `&`, `or`, or `and`. This should, if done correctly, result in a column of featured artist(s) that sometimes has one artist, and sometimes has a character vector of multiple artists. 
    
        You could then drop any songs with no featued artist(s) since we don't need to worry about them any more. Pull out the featured artist(s) column and, once again, use `unlist()` again to turn the result into a vector where each element is a single artist's name. Use `str_trim()` to remove any annoying whitespace at the start or end of each element, put the rest back into tibble form, then count the number of times each artist occurs. Print the first however many highest values you'd like.
        
        If you've done this correctly, the second-most common featued artist should be "DRAKE" with 40 credits.
    
**To submit this assignment:**

Ideally, knit straight to PDF by changing `html_document` to `pdf_document` in line 5 above. This should work as long as you properly installed LaTeX in Tutorial 0.1. Otherwise:

1. Knit to HTML. An HTML document should open automatically in another RStudio window.

2. Click "Open in Browser" in that HTML document. It should open as a webpage in your default browser (e.g. Chrome).

3. Click Ctrl+P/Command+P, but instead of printing a hard copy on your printer click "Save as PDF."

4. Save and upload that document to Canvas.
    
    


------BEGIN ANSWER BELOW--------

## Load necessary packages

```{r load raw_ab}
raw_ab <- as_tibble(read.csv("./Datasets/raw_ab.csv"))
# raw_ab <- read.csv("./Datasets/raw_ab.csv")
```

# Dr. Martin van der Linden HW5 : Exercises with `stringr`

We will be working with Dr. Martin van der Linden's HW5 assignment. Specifically, we will clean the data in `raw_ab.csv` - a dataset containing abortion data from both the **Guttmacher Institue** and the **CDC**. 

## Removing Missing Values from `raw_ab`

We will start by removing all rows that either:

* Does not contain any number
* is `NA`

We will accomplish that by using `str_detect()` in conjunction with regex to keep only rows that have digits in them. Since rows containing the value `NA` do not have any digits, this operation also removes rows with `NA`. We then store this new dataframe as `ab1`.

```{r remove rows where CDC is empty or NA}

ab1 <- raw_ab[str_detect(raw_ab$CDC, "\\d"),] %>% 
  drop_na("CDC")
ab1
```

## Removing Parentheses from the `CDC` and `AGI` columns of `ab1`

Some of the values in the `CDC` and `AGI` columns in `ab1` contain parentheses. If we are to transform the columns into numerical data for computation, we must remove these parentheses. We use the string escape character with the regex escape character as well as the "or" (`|`) to find all parentheses, which we then replace with empty strings. This effectively removes all parentheses.
```{r remove parentheses}
ab2 <- ab1
ab2$CDC <- ab2$CDC %>% 
  str_replace_all("\\(|\\)", "")
ab2$AGI <- ab2$AGI %>% 
  str_replace_all("\\(|\\)", "")
ab2
```

## Removing Thousand Separators in the `CDC` and `AGI` columns of `ab2`

The values in the `CDCD` and `AGI` columns, in their raw form, are formatted such that every thousand, or 3 digits starting from the right, there is a `,` acting as a separator. Normally, this would increase human readability of large numbers, but in order to do numerical operations on these numbers we have to remove the commas. We do an operation similar to what we did with the parenthesis - finding commas with regex and replacing them with empty strings.

```{r remove commas}
ab3 <- ab2
ab3$CDC <- ab3$CDC %>% 
  str_replace_all(",", "")
ab3$AGI <- ab3$AGI %>% 
  str_replace_all(",", "")
ab3
```

## Converting `ab3` to Numeric

We are now ready to convert our numeric values to numeric types in R, which allows for computations. Using `select()`, we first drop the column `X` - we will not need this column. We will then use array indexing to grab the columns we want, then use `as.numeric()` in conjunction with `apply()` to execute this conversion.

```{r numericalize}
ab4 <- ab3 %>% 
  select("year":"state")
ab4[1:3] <- apply(ab4[,1:3], 2, function(x) as.numeric(x))
ab4
```

# Decoding the Secret Message in `Puzzle 4.1`

Apparently there is a secret message hidden in [this puzzle]("https://www.adventures-in-r.com/pas/pa-regex"). Let's load it and see what it is about. 

```{r load message}
message <- read.csv("https://www.dropbox.com/s/lgpn3vmksk3ssdo/scrambled_message.txt?dl=1", stringsAsFactors = FALSE)$Word
head(message)
```

It looks like we have a very messy character vector. We will first do the Warm-up exercises, then try to decode the message.

## Warm-up Exercises

### Number of Characters
The first warm-up wants us to find the number of characters there are in the message. We will assume that this means removing the spaces with `str_trim()` then summing the lengths (`str_length()`) of the remaining strings.

```{r number of characters}
message %>% 
  str_trim() %>% 
  str_length() %>% 
  sum()
```

### Number of Words
According to the challenge, a “word” is a set of characters with no white space - in other words, any non-whitespace element in the `message` vector is a "word". We will use `str_trim()` to remove the whitespace then use `length()` to find the number of elements in the vector.
```{r number of words}
message %>% 
  str_trim() %>% 
  length()
```

### Show Punctuation Marks
The next warm-up exercise wants us to show all the punctuation marks in the scrambled message. To view all matches and highlight punctuation, we employ `str_view_all()`, and use the regex character class `[:punct:]` to match all punctuation.
```{r show punctuation}
str_view_all(message, "[:punct:]")
```

### Printing Longest Word in Capitals
The final recommended warm-up exercise wants us to print the longest "word" in the scrambled message in all caps. To do this, we must first identify the longest word in the message. We can do that with a composition of `str_length()` and `max()`. Now that we know the max length, we can use this length as a condition to match the string with this length in `message` by using array indexing `[]`. Finally, we use `str_to_upper()` to capitalize the "word".
```{r print longest word in caps}
str_to_upper(message[str_length(message) == max(str_length(message))])
```

## Decoding the Message

We will now follow the steps on the challenge page to decode the message. We first remove any whitespace before and after each word using `str_trim()`.
```{r remove whitespace}
message <- message %>% 
  str_trim()
```

Then we drop all characters off the end of each word that is more than 16 characters long. To do so, we first locate all strings longer than 16 characters, then we replace those strings with their respective first 16 characters.
```{r make length 16}
message[str_length(message) > 16] <- str_sub(message[str_length(message) > 16], 1, 16)

# Debug
max(str_length(message))
```

